---
title: "Lab 6: Kaplan Meier Survival Analysis"
output: html_document
---  

### Lab Learning Objectives  
- To estimate survival function and median survival time 
- Plot *Kaplan-Meier* survival curve using SAS 
- Statistically compare survival curves  

***  

### Dataset  
- `hivhepc.sas7bdat` - <span style="color:purple">This dataset will also be used for Lab 7</span> 
- Cohort study on human immunodeficiency virus (HIV) infection 
- Includes medical, behavioral, and epidemiologic information from HIV positive individuals 
- *Research question*: Whether having a diagnosis of hepatitis C is associated with a shorter median survival time?  

***  

### Data dictionary of the HIVHepC dataset  

Variable | Description | Range |  
--- | --- | --- |  
patient | Identifier for the Amsterdam Cohort Studies participant | Random number |  
MSM | Men who have sex with men | 0 = No <br> 1 = Yes |  
IVDrugUse  | History of drug use | 0 = No <br> 1 = Yes |  
ART_init | Has ART been initiated? | 0 = No <br> 1 = Yes |  
HepC | Is the patient HepC positive? | 0 = No <br> 1 = Yes |  
cohort_date | Date of enrollment into cohort anonymous <br> (changed to 1st day of month) - HIV diagnosis | dd-mm-yyyy |  
death | Has the subject died? | 0 = No <br> 1 = Yes |  
last_alive_date | Date last alive anonymous <br> (changed to 1st day of month) | dd-mm-yyyy |  
death_date | Date of death anonymous <br> (changed to 1st day of month) | dd-mm-yyyy |  
age | Age of patient at the last clinical visit | 99 (years) |  
HIV_diag_1996 | Was HIV diagnosed before 1996? | 0 = No <br> 1 = Yes |  

***  

##### Copy the `HIVHepC.sas7bdat` from Courseworks to your flash drives or desktop.  

# {.tabset .tabset-fade}  

## SAS  
In the `libname` statement, the `...` denote the path to your folder with HIVHepC data.  

```{}
libname Epi3 '...’;
data hivhepc;
set Epi3.hivhepc;
run;
```

## R  

First, we will need to load all necessary packages for our data analysis.  
```{r load packages, message = FALSE}
library(tidyverse) # For data cleaning and tidying  
library(haven) # For importing SAS dataset(s)  
library(survival) # For our Kaplan Meier survival analysis
library(survminer) # For K-M survival analysis plotting  
```

Next, we will import the `hivhepc` dataset and 'tidy' the variable names.  
```{r import dataset}
hivhepc <- read_sas('./datasets/hivhepc.sas7bdat') %>% 
    janitor::clean_names()
```

We can see from our `Global Environment` that the import was successful.  

# {-}

***  

#### 1. How many observations are in the data?  

# {.tabset .tabset-fade}  

## SAS  
```{}
proc contents data = hivhepc; run;
```

<details>
<summary>*Show answer*</summary> 
<p>

<span style="color:blue">There are 329 observations and 11 variables.</span>  

</details>  

## R  

```{r}
# Number of observations (or rows) 
nrow(hivhepc)

# Number of variables (or columns) 
ncol(hivhepc)
```

<details>
<summary>*Show answer*</summary> 
<p>

<span style="color:blue">There are `r nrow(hivhepc)` observations and `r ncol(hivhepc)` variables.</span>  

</details>  

# {-}  

***  

#### 2. How many MSM and how many intravenous drug users? How many have both risk factors? How many have neither?  

# {.tabset .tabset-fade}  

## SAS  
```{}
proc freq data= hivhepc;
table MSM IVdruguse MSM*IVDrugUse HepC / list; 
run;
```

## R  
We will need to generate our frequency tables first. We can do this using the `tabyl()` function as part of the `janitor` package.  

```{r sample frequencies}
# One-way frequency tables -> tibbles
msm_count <- hivhepc %>% janitor::tabyl(msm) %>% as_tibble()
ivdruguser_count <- hivhepc %>% janitor::tabyl(ivdruguse) %>% as_tibble()

# Two-way frequency table (using `xtabs`)
msm_and_ivdu_count <- xtabs(~ msm + ivdruguse, data = hivhepc)
```

Now, we can print our generated frequency tables.  

```{r print sample frequencies}
# Print frequency tables 
msm_count
ivdruguser_count
msm_and_ivdu_count
```

Recall from the data dictionary that `msm = 1` indicates that the participant *has sex with men*, and that `ivdruguse = 1` indicates that the participant has a *history of iv drug use*. Knowing this, we can see from our tables that:  

- there are `r msm_count %>% filter(msm == 1) %>% pull(n)` MSM  
- there are `r ivdruguser_count %>% filter(ivdruguse == 1) %>% pull(n)` participants with a history of IV drug use 
- there are `r msm_and_ivdu_count[[2,2]]` participants that are both MSM and participants with a history of IV drug use 
- there are `r msm_and_ivdu_count[[1,2]]` participants that have neither risk factor  

# {-}  

***  

#### 3. How many of the participants have had a diagnosis of Hepatitis C?  

The table below indicates how many and what percentage of our sample have had a diagnosis of Hepatitis C.  

```{r}
hivhepc %>% janitor::tabyl(hepc) %>% filter(hepc == 1)
```

***  

#### 4. In a cohort study, what are the four events that would determine when an individual stops contributing person-time?  
<details>
<summary>*Show answer*</summary> 
<p>

<span style="color:blue">– Person dies  
– Person develops the disease of interest  
– Person is lost to follow-up  
– Study period ends</span>  

</details>  

***  

#### 5. Is this is a closed or open cohort? How would we examine the data to determine this?  

<details>
<summary>*Show answer*</summary> 
<p>

<span style="color:blue">– Check to see if the follow up time is the same for all subjects that did not die  
– The enrollment date (cohort_date) and the last follow-up date (last_alive_date) would be same  
– If it were a closed cohort, there would be no losses to follow up</span>  

<span style="color:blue">Note that it is possible to have some variability in enrollment date (because it’s not always possible to enroll large cohorts in a single day!) but in general, closed cohorts have a single enrollment event/period, and the same length of follow-up for all those that do not have the event of interest.</span>  

</details>  

***

# {.tabset .tabset-fade}  

## SAS  
```{}
/*Question 5*/
proc freq data = hivhepc;
table cohort_date*last_alive_date / list;
where death = 0;
run;
```

You can see from a sample of the `PROC FREQ` output that `cohort_date` varies for participants. In addition, the `last_alive_date` is different for all participants. In a closed cohort we would expect that all participants would have the same cohort start date and the same last_alive_date (for participants who did not die).  

## R  
```{r}
# Code to be entered later  
```

# {-}

***  

#### 6. The amount of time each member of the cohort is followed is not provided for you. Using the variables provided how would you calculate it?  

<details>
<summary>*Show answer*</summary> 
<p>

<span style="color:blue">- Among people who died, subtract enrollment date (cohort_date) from date of death  
- Among people who did not die, subtract enrollment date from last date alive</span>  

<span style="color:blue">Note: To be able to subtract dates to get the days of follow- up, the dates must be in the same format, e.g. MMDDYYYY, MMYYYY, etc.</span>  

</details>  

# {.tabset .tabset-fade}  

## SAS  

Calculate person year using SAS  

```{}
data hivhepc2;
set hivhepc;

if death = 1 then days = death_date - cohort_date;
else if death = 0 then days = last_alive_date - cohort_date;
pyears = days/365.25; *gives time-to-event in years;

run; 
```

<span style="color:blue">We convert to years for best interpretability, but other units could be used.</span>  

Check your work:  

```{}
proc freq data = hivhepc2;
table death_date*cohort_date*pyears / list missing nocum nofreq nopercent;
run;

proc freq data = hivhepc2;
table last_alive_date*cohort_date*pyears / list missing nocum nofreq nopercent;
run;

proc freq data = hivhepc2;
table pyears;
run;
```

## R  
```{r}
# Code to be entered later  
```

# {-}  

***  

### Part II  

***  

#### 1. How many people had died at the end of follow-up? How many people were censored (did not have the event of interest)?  
# {.tabset .tabset-fade}  

## SAS  
```{}
proc lifetest 
    data = hivhepc2
    method = km
    plots = (s)
    graphics; 
    time pyears*death(0);
run;
```

<span style="color:blue">`pyears` is the time variable <br> `death` is the event (`Death = 1`) <br> `(0)` is the censored value (`Death = 0`)</span> 

## R  
```{r}
# Code to be entered later  
```

# {-}  

<details>
<summary>*Show answer*</summary> 
<p>

<span style="color:blue">Patients died during follow-up: 194 <br> Patients censored: 135</span>  

</details>  

***  

#### 2. Examine the survival function using Kaplan-Meier survival curve of the entire study population. Using the plot to visually estimate, what is the median overall survival time?  

<details>
<summary>*Show answer*</summary> 
<p>

<span style="color:blue">Median survival time: Between 12 and 12.5 years <br> (time corresponding to a survival function estimate of 0.5)</span>  

<span style="color:blue">Tracing a line from the survival function = 0.5 (Y axis), and the time corresponding to it on the X axis.</span>

</details>  

***  

#### 3. Using the output, what is the median survival time?  

<details>
<summary>*Show answer*</summary> 
<p>

<span style="color:blue">Median survival time is 12.08 years</span>  

<span style="color:blue">The log-log transformation is the default method in SAS for calculating CIs for survival analysis. The confidence limits are calculated by adding and subtracting 1.96 x SE to the log(-log) transformation of the survivor function, or log(-logS(t)) and then reversing the transformation to get back to the original scale. **This method is preferred because it ensures that the confidence limits will not be greater than 1 or less than 0.**</span>  

</details>  

***  

#### 4. What is the probability of surviving beyond 5 years from the time of HIV diagnosis?  

